Structuring Theories with Implicit Morphisms

### 2018-09-27 mmt.tex
update

@@

@@

\begin{example}[Includes]\label{syn:incl}
We extend the grammar with
\begin{grammar}
  Dec & \icl{S} & \text{include a theory into a theory} \\
  Ass & \icl{M} & \text{include a morphism into a morphism}
\end{grammar}
\end{example}

>>>>

\begin{example}\label{ex:thgroup}
The (flat) theory for $\cn{Group}$ and $\cn{DivGroup}$ as presented in the introduction are:
\begin{mmtcode}
theory Group : ?Meta =
  universe	: type		❘ # U ❙
  op		: U ⟶ U ⟶ U	❘ # 1 ∘ 2 ❙
  unit		: U		❘ # e❙
  inverse : U ⟶ U ❘  # 1 ⁻¹ ❙

  axiom_assoc		: ⊦ ∀[x]∀[y]∀[z] x ∘ (y ∘ z) ≐ (x ∘ y) ∘ z ❙
  axiom_right_unit	: ⊦ ∀[x] x ∘ e ≐ x ❙
  axiom_left_unit	: ⊦ ∀[x] e ∘ x ≐ x ❙
  axiom_inverse : ⊦ ∀[x] x ∘ (x⁻¹) ≐ e ❙
❚
\end{mmtcode}
\begin{mmtcode}
theory DivisionGroup : ?Meta =
  universe : type ❘ # U prec 1 ❙
  div : U ⟶ U ⟶ U ❘ # 1 / 2 ❙
  nonEmpty : U ❘ # e prec 1 ❙
  // axioms ... ❙
❚
\end{mmtcode}
where $\GS,\US,\RS$ are delimiters and the (optional) \lstinline|# <not>| is an optional notation. \lstinline|?Meta| is an intended meta-theory containing e.g. the symbols $\forall,\doteq$ etc.
\end{example}

@@

\begin{example}
	We can now specify the morphisms between our two definitions of groups. The theory of groups as monoids with inverses is given in \autoref{syn:incl}, the theory $\cn{DivGroup}$ and the morphism $\cn{DG2G}$ are as follows:
\begin{mmtcode}
view DG2G : ?DivisionGroup -> ?Group =
  universe = universe ❙
  div = [a,b] a ∘ (b⁻¹) ❙
  nonEmpty = e ❙
  // axioms ...❙
❚
\end{mmtcode}
The morphism $\cn{DG2G}$ hence maps the universe of a division group to the universe of a group, division to the expression $\lambda a,b.\; a\circ b^{-1}$ and the witness $\cn{nonEmpty}$ of a division group to the unit $e$ of a group. Additionally, the view has to map the axioms of a division group to their \emph{proofs} in $\cn{Group}$ of their translated statements.

The reverse morphism $\cn{G2DG}$ is completely analogous, mapping $\cn{op}$, $\cn{unit}$ and $\cn{inverse}$ to their corresponding definitions in terms of the division operation $\cn{div}$.
\end{example}

@@

\begin{example}[Includes]\label{syn:incl}
We extend the grammar with
\begin{grammar}
  Dec & \icl{S} & \text{include a theory into a theory} \\
  Ass & \icl{M} & \text{include a morphism into a morphism}
\end{grammar}
This allows writing the theory $\cn{Group}$ as in \autoref{ex:thgroup} by extending a theory of monoids:
\begin{mmtcode}
theory Monoid : ?Meta =
  universe	: type		❘ # U ❙
  op		: U ⟶ U ⟶ U	❘ # 1 ∘ 2 ❙
  unit		: U		❘ # e❙

  axiom_assoc		: ⊦ ∀[x]∀[y]∀[z] x ∘ (y ∘ z) ≐ (x ∘ y) ∘ z ❙
  axiom_right_unit	: ⊦ ∀[x] x ∘ e ≐ x ❙
  axiom_left_unit	: ⊦ ∀[x] e ∘ x ≐ x ❙
❚

theory Group : ?Meta =
  include ?Monoid ❙

  inverse : U ⟶ U ❘  # 1 ⁻¹ ❙
  axiom_inverse : ⊦ ∀[x] x ∘ (x⁻¹) ≐ e ❙
❚
\end{mmtcode}
The flat variant $\cn{Group}^\flat$ we get by making all included symbols (from $\cn{Monoid}$) available to $\cn{Group}$, resulting in exactly the theory as presented earlier in \autoref{ex:thgroup}.
\end{example}
